---
title: Data Analysis Project Proposal
subtitle: Proposal
author: "Srilakshmi Venkatesan, Sriya Sridhar, Ryan Lee, Daniel Murong"
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

-   Address ethical concerns about the data.
    If none are identified, state that explicitly.

## Research question

-   A well formulated research question.
    (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    What is the trend of delays/cancellations throughout the year and what are common predictors of those?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    The research topic focuses on how the number of delays/cancellations change throughout the year depending on the month.
    In addition, it focuses on the causation behind the delays/cancellations.
    We predict that delays/cancellations are greater around the winter months due to the weather, and around the holiday season due to the volume of travelers.

-   Identify the types of variables in your research question.
    Categorical?
    Numerical?

    The categorical variables in our research question are the predictors of cancellations, such as weather and year/month.
    The numerical variables in our research question are the number of delays due to factors such as weather or security.

## Glimpse of data

```{r}
#| label: load-data-1

# add code here
```

# Data 2

## Introduction and data

The source of the data is the UC Irvine Machine Learning Repository.
The data was collected through school reports and questionnaires by Cortez and Silva for a study published in April 2008.
Each observation is collected from student and then describing their age, sex, address, job, health, etc.
No ethical concerns stated regarding the data and none to be found.

## Research question

<<<<<<< HEAD
Do demographical factors such as age, sex, home, parental status, and parental education impact student performance in math in the Portugueuse schools of Gabriel Pereira and Mousinho da Silveira? Or are factors such as study time, school travel time, educational support, and paid class time more influential?
=======
-   A well formulated research question.
    (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)
>>>>>>> 5dfb937d0bf9348817f65859eccb71eef5bb16dd

The research topic is what topics are most influential in educational performance on math, home characteristics or study time and study resources. There may be some overlap between the two, as certain factors may lead to more or less study time. Therefore, we also want to explore the relationship between the two. Our hypothesis is that study time and educational support are more influential n mathematic educational performance. 

<<<<<<< HEAD
The types of variable present include numerical variables such as age, study time, school travel time, paid class versus categorical and binary variables in parental status, home, sex, parental education, and educational support. 
=======
-   Identify the types of variables in your research question.
    Categorical?
    Numerical?
>>>>>>> 5dfb937d0bf9348817f65859eccb71eef5bb16dd

## Glimpse of data

```{r}
#| label: load-data-2

# add code here
```
