---
title: Data Analysis Project Proposal
subtitle: Proposal
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
library(readxl)
```

# Data 1

## Introduction and data

  The airlines dataset is from the CORGIS Dataset Project. The data was originally collected by the Bureau of Transportation Statistics. This dataset was curated by Austin Cory Bart. Each observation is on an airport on a given date with variables including the number of delays and cancellations attributed to a certain event, such as meteorological conditions. No ethical concerns are identified for this dataset.

## Research question

  What is the trend of delays/cancellations throughout the year and what are common predictors of those?

  The research topic focuses on how the number of delays/cancellations change throughout the year depending on the month. In addition, it focuses on the causation behind the delays/cancellations.
    We predict that delays/cancellations are greater around the winter months due to the weather, and around the holiday season due to the volume of travelers.

  The categorical variables in our research question are the predictors of cancellations, such as weather and year/month.
    The numerical variables in our research question are the number of delays due to factors such as weather or security.

## Glimpse of data

```{r}
#| label: load-data-1
airlines <- read_csv("data/airlines.csv")
glimpse(airlines)
```

# Data 2

## Introduction and data

  The source of the data is the UC Irvine Machine Learning Repository. The data was collected through school reports and questionnaires by Cortez and Silva for a study published in April 2008. Each observation is collected from student and then describing their age, sex, address, job, health, etc. No ethical concerns stated regarding the data and none to be found.

## Research question
    
  Do demographical factors such as age, sex, home, parental status, and parental education impact student performance in math in the Portugueuse schools of Gabriel Pereira and Mousinho da Silveira? Or are factors such as study time, school travel time, educational support, and paid class time more influential?
    
  The research topic is what topics are most influential in educational performance on math, home characteristics or study time and study resources. There may be some overlap between the two, as certain factors may lead to more or less study time. Therefore, we also want to explore the relationship between the two. Our hypothesis is that study time and educational support are more influential in mathematic educational performance.
    
  The types of variable present include numerical variables such as age, study time, school travel time, paid class versus categorical and binary variables in parental status, home, sex, parental education, and educational support.

## Glimpse of data

```{r}
#| label: load-data-2
edu_achievement <- read_excel(
  "data/educationachievement.xlsx",
  skip = 1
)
glimpse(edu_achievement)
```
