---
title: Data Analysis Project Proposal
subtitle: Proposal
author: "Srilakshmi Venkatesan, Sriya Sridhar, Ryan Lee, Daniel Murong"
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

-   Address ethical concerns about the data.
    If none are identified, state that explicitly.

## Research question

-   A well formulated research question.
    (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    What is the trend of delays/cancellations throughout the year and what are common predictors of those?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    The research topic focuses on how the number of delays/cancellations change throughout the year depending on the month.
    In addition, it focuses on the causation behind the delays/cancellations.
    We predict that delays/cancellations are greater around the winter months due to the weather, and around the holiday season due to the volume of travelers.

-   Identify the types of variables in your research question.
    Categorical?
    Numerical?

    The categorical variables in our research question are the predictors of cancellations, such as weather and year/month.
    The numerical variables in our research question are the number of delays due to factors such as weather or security.

## Glimpse of data

```{r}
#| label: load-data-1

# add code here
```

# Data 2

## Introduction and data

The source of the data is the UC Irvine Machine Learning Repository.
The data was collected through school reports and questionnaires by Cortez and Silva for a study published in April 2008.
Each observation is collected from student and then describing their age, sex, address, job, health, etc.
No ethical concerns stated regarding the data and none to be found.

## Research question

-   A well formulated research question.
    (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question.
    Categorical?
    Numerical?

## Glimpse of data

```{r}
#| label: load-data-2

# add code here
```
